---
title: "TBD"
author: "Tomaso Bulligan"
date: "Tuesday, September 16, 2014"
output: html_document
---

```{r options,echo=FALSE}
library(knitr)
library(data.table)
opts_chunk$set(dev='CairoPNG')
setwd("./GitHub/RepData_PeerAssessment2/")
```

# Notes

REMEMBER TO REMOVE SETWD

## Instructions
https://class.coursera.org/repdata-006/human_grading/view/courses/972592/assessments/4/submissions

## Important *codebook like* resources
https://ire.org/nicar/database-library/databases/storm-events/
http://www.ncdc.noaa.gov/stormevents/

## Questions

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, **there is no need to make any specific recommendations in your report**.

# !!! TITLE TBD !!!

# Synopsis

# Data Processing

```{r}
# Fetch original data set
file.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
file.local <- "StormData.csv.bz2"
if(!file.exists(file.local)) {
    download.file(file.url, file.local, "curl")
}

# Read data set - WHY DATA.TABLE DOESN'T WORK?
df <- read.csv("StormData.csv.bz2", stringsAsFactors = FALSE)

# Clean data previous to 1996 (explain why)

# Consider cleaning people/property/crop damage == 0
```

# Results

# 